Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2023-01-07 edubots_surveys.Rnw

%%% Load packages
\usepackage{amsthm,amsmath}
\usepackage{multirow}
\usepackage{rotating}
\usepackage[utf8]{inputenc} %unicode support

@@

\textcolor{blue}{In recent years, new technologies are being widely adopted and introduced into the classroom. However, in order to be successful, their application 
will require additional work and training for the teacher, as well as for the students. Thus, both of them will acquire new skills and, potentially, pupils will be more 
engaged to the subject, which is usually the main objective. Even if several technologies are already in use in the classroom, there is still room for adopting new others.
Anyway, this adoption strongly depends on the compromise and capabilities of the teachers, as main active actors in this new academic approach. Asking and analyzing users' 
opinions with regard to this scenario will help reaping the expected benefits of the technology}.

\textcolor{blue}{Messaging applications such as WhatsApp or Telegram are becoming very popular in teaching scenarios mainly at secondary and higher education.
In addition to their main use, some of these tools offer the possibility to interact with \textit{chatbots} (or just bots), even supporting synchronous conversations between the different
parties \cite{differ.chat}. It is also possible to create a standalone chatbot application and attach it to a messaging tool through an API (Application Programming Interface).}

\textcolor{blue}{\textit{Chatbots} are autonomous agents that are able to communicate and interact in a human-like way \cite{Gong2008}, answering question and even posing their own ones to the user. 
Thus, we can say that they are close to passing a Turing Test \cite{moor2003turing}. They are implemented by means of different Artificial Intelligence techniques, such as natural language processing.
Within the plethora of possible applications of chatbots, education is one of the most prolific nowadays \cite{clarizia2018,Smutny2020}. In this domain, they facilitate the so-called 
\textit{personalized learning}, i.e. they adapt themselves to students' self-pace learning and provide online tutoring. Personalized learning has been proved to be very effective increasing students course engagement and participation \cite{agarwal2020}. Moreover, chatbots also reduce students' stress \cite{Daniel2016,agarwal2020} by avoiding the face-to-face tutoring with their teachers, 
or asking questions in front of the class. This situation usually results in a later contact with the teacher by e-mail, which in turn, will mean a workload for the teacher that could be very 
significant given the disproportionate ratio of number of students to teachers.  
}
Hence, the chatbot technology has a potential to mitigate this problem by providing answers to students’ questions and facilitating a dynamic and autonomous learning
experience \cite{Griol2014,kim2020}. In addition, using an automated system such as a chatbot will allow teachers to spend more time on  other topics that students struggle with.

Given the potential benefits that chatbots could bring to the classroom, the
focus of the % EDUBOTS KA2 European project \footnote{https://www.edubots.eu},
unnamed European project, the main sponsor of this study, is to explore the best practices and innovative use of chatbots, 
and to create a learning community of educators in higher education institutions.
\textcolor{blue}{Researchers and teachers participating in this project have developed several studies. Most of these works are focused on how  
messaging applications and/or chatbots are normally applied to deliver personalized learning in classrooms that occurs anytime anywhere,
promote collaborative learning experiences, group discussions \cite{panah2020study} and boost students' sense of belonging to their institutions
\cite{abbas2021onlinechat}.}
\textcolor{blue}{The majority of these studies were conducted focusing on the teachers' perspectives, so this paper aims to provide more conclusions regarding 
their opinions and needs, as well as the existing challenges or opportunities that arise when adopting these technologies in classrooms. This will lead to clarify
how they can positively contribute to enhancing the learning process in higher education institutions.}

In this study, data collection involved two phases:
 \begin{itemize}
\item The first phase aimed to collect feedback from students about their
  students’ preferences of tools/applications for chatting and
  messaging, how they use them in educational context, who they like to be
  with in the class messaging groups, and their expectations of chatbots 
  in assisting them during their learning process. Therefore, two surveys
  for bachelor and master degree students at the University of Granada (Spain)
  were designed and answers from more than 250 students were collected. The key
  findings of the students' surveys reveal that they prefer to use messaging
  familiar applications, such as Telegram or WhatsApp. 
  Students also expressed interest in chatbots that can assist them with
  organizing their course schedules, help them access their assignment grades
  and facilitate searching for resources.
\item The results of the first phase were fed into the second phase of the
  study, which focused on teachers’ opinions about using chatbots and messaging
  applications in classes. These results are the main focus of this paper.
\end{itemize}

Initially, a single survey was developed for this purpose and responses from 300
higher education teachers were collected and analysed\textcolor{black}{; this will be detailed in Section \ref{sec:meth}}. The responses' analysis
led to the preparation of a second survey for teachers and 200 responses were
collected\textcolor{black}{, again following the methodology that will be explained later. The questions of the second survey focused on asking about teachers'
needs, opinions and preferences of the development of future technology-enhanced
tools and its potential impact on educational institutions policies. In order to assess these, this paper layed out the following research questions}:

\begin{itemize}
\item RQ1 - Do teachers use now or want to use messaging apps in their classes?
\item RQ2 - Which chatbots' features would teachers find useful for this purpose?
\item RQ3 - Which kind of interaction do teachers prefer with their students?
\item RQ4 - Which kind of interaction media features do teachers value the most?
\end{itemize}

@@

%A total of 205 teachers responded to the second survey, 124 males (60.5\%), 65 females (31.7\%) and 16 individuals who chose not to identify their gender (7.8\%). The majority of the teachers were aged 45-55 years old (70; 34.1\%), followed by individuals aged 35-45 (67, 32.7\%), and 25-35 (42; 20.5\%), while the minority was aged $>$~55 years old (26; 12.7\%). In terms of teaching experience, 59 teachers (28.8\%) had an experience of 16-25 years, followed by teachers who had a teaching experience of 06-15 years (51; 24.9\%), 0-5 years (51; 24.9\%) and more than 25 years of experience (44; 21.5\%). The teachers were asked about what kind of interaction media features they value the most \textcolor{black}{; i.e. what their preferred features are in the applications/tools that they use to interact with their students.} The frequencies and percentages of their responses are given in Table \ref{tab:interaction}.

\begin{table}[h!]
\begin{tabular}{lcccc}
\hline
Interaction features & \multicolumn{2}{c}{Yes} & \multicolumn{2}{c}{No} \\
\hline
                    &  Freq. & \% &  Freq. & \% \\
                    \cline{2-5}

Analytics & 108 & 52.7 & 97 & 47.3 \\
Connectivity & 119 & 58.0 & 86 & 42.0 \\
Familiarity & 121 & 59.0 & 84 & 41.0 \\
Hidden Phone Number & 113 & 55.1 & 92 & 44.9 \\
Horizontally & 134 & 65.4 & 71 & 34.6 \\
Official formation & 42 & 20.5 & 163 & 79.5 \\
Pluggability & 65 & 31.7 & 140 & 68.3 \\
Sustainability & 157 & 76.6 & 48 & 23.4\\
Unidirectionality & 27 & 13.2 & 178 & 86.8 \\
Officiality & 150 & 73.2 & 55 & 26.8 \\
Synchrony & 45 & 22.0 & 160 & 78.0\\
Other & 5 & 2.4 & 200 & 97.6 \\
\hline
\end{tabular}
\vspace{0.8em}
\caption{Interaction media features valued by teachers. Details about each of these features can be read in Appendix I (Section \ref{app:questions}, second survey).}
\label{tab:interaction}
\end{table}

@@

The datasets generated and analysed during the current study are available in the {\tt anonymous} repository, accessible from https://anonymous.URL
% {\tt https://github.com/JJ/edubots-paper/tree/main/data}.

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
\usepackage{multirow}
\usepackage{rotating}
\usepackage[utf8]{inputenc} %unicode support
\usepackage{hyperref} % for URL

@@

\textcolor{blue}{New technologies are being introduced and widely adopted into the classroom in the last years. However, in order to be successful, their application 
will require (many) extra work and training for the teacher, as well as for the students. Thus, both of them will acquire new skills and, potentially, pupils will 
get a higher engagement to the subject, which is usually the main objective. Even if several technologies are already in use inside the classroom, there is still room for adopting others.
Anyway, this adoption strongly depends on the compromise and capabilities of the teachers, as main active actors in this new academic approach. Given this, 
asking and analyzing user's opinions with regard to this scenario will help reaping the expected benefits of the technology}.

\textcolor{blue}{In the last years, messaging applications such as WhatsApp or Telegram are being used in teaching scenarios mainly at secondary and higher education.
In addition to their main use, some of these tools offer the possibility of interact with \textit{chatbots} (or just bots), even supporting synchronous conversation between the different
parties \cite{differ.chat}. There is also possible to create a standalone chatbot application and attach it to a messaging tool through an API (Application Program Interface).}

\textcolor{blue}{\textit{Chatbots} are autonomous agents able to communicate and interact with humans (or other chatbots) in a human-like way \cite{Gong2008}, answering question and even posing their own ones to the user. 
Thus, we can say that they are close to passing a Turing Test \cite{moor2003turing}. They are implemented by means of different Artificial Intelligence techniques, such as natural language processing.
Inside the plethora of possible applications of chatbots, education is one of the most prolific nowadays \cite{clarizia2018,Smutny2020}. In this domain, they facilitate the so-called 
\textit{personalized learning}, i.e. they adapt themselves to students’ pace of learning and provide online tutoring outside the classroom, which have been proved to be very effective increasing their 
engagement and participation in the subjects \cite{agarwal2020}. Moreover chatbots' tutoring possibilities aid to reduce the stress of students \cite{Daniel2016,agarwal2020} by avoiding the face-to-face tutoring with their teacher, 
or asking questions in front of the class. This situation normally conducts to a silence of the student and a later contact with the teacher by e-mail, which, in turn, will mean a workload for the teacher, that could be very 
significant given the existing disproportion between number of students per each teacher.  
}
Hence, the chatbot technology has a potential to mitigate this problem by providing answers to students’ questions and facilitating a dynamic and autonomous learning
experience \cite{Griol2014,kim2020}. In addition, using an automated system such as a chatbot, can draw teachers’ attention to topics that students struggle to comprehend or need further assistance in understanding.

Given the potential benefits that chatbots could bring to the classroom, the
focus of the % EDUBOTS KA2 European project \footnote{https://www.edubots.eu},
unnamed European project, the main sponsor of this study, is to explore best practices and innovative use of chatbots, 
and to create a learning community of educators in higher education institutions.
\textcolor{blue}{Researchers and teachers participating in this project have developed several studies. Most of these works have been focused on how the 
messaging applications and/or chatbots are normally applied to deliver personalized learning in classrooms that occurs anytime anywhere,
promote collaborative learning experiences, group discussions \cite{panah2020study} and boost students' sense of belonging to their institutions
\cite{abbas2021onlinechat}.}
\textcolor{blue}{The majority of these studies were conducted focusing on the teachers' perspectives, so this paper aims to provide more conclusions regarding 
their opinions and needs, as well as the existing challenges or opportunities that arise when adopting these technologies in classrooms. This will lead to clarify
how they can positively contribute to enhancing the learning process in higher education institutions.}

In this study, data collection involved two phases:
 \begin{itemize}
\item The first phase aimed to collect feedback from students about their actual
  information about students’ preferences of tools/applications for chatting and
  messaging, how they use them in educational context, who do they like to be
  with in the class messaging groups, and their expectations of the chatbots in
  terms of assisting them during their learning process. Therefore, two surveys
  for bachelor and master degree students at the University of Granada (Spain)
  were designed and answers from more than 250 students were collected. The key
  findings of the students' surveys reveal that they prefer to use messaging
  applications, such as Telegram or WhatsApp as they are familiar to
  them. Students also expressed interest in chatbots that can assist them with
  organizing their course schedules, help them access their assignment grades
  and facilitate searching for course's resources.
\item The results of the first phase were fed into the second phase of the
  study, which focused on teachers’ opinions about using chatbots and messaging
  applications in classes. This is what is going to be presented in this paper.
\end{itemize}

Initially, a single survey was developed for this purpose and responses from 300
higher education teachers were collected and analysed\textcolor{black}{; this will be explained more fully in the methodology section}. The responses' analysis
entailed the creation of a second survey to which 200 teachers' responses were
collected\textcolor{black}{, with the methodology and collective characteristics that will be explained below. The questions of the second survey focused on discussing teachers'
needs, opinions and preferences of the development of future technology-enhanced
tools and its potential impact on educational institutions’ policies. In order to assess these, this paper investigated the following research questions}:

\begin{itemize}
\item RQ1 - Do teachers use now or want to use messaging apps in their classes?
\item RQ2 - Which chatbots' features would teachers find useful for this purpose?
\item RQ3 - Which kind of interaction do teachers prefer with their students?
\item RQ4 - What kind of interaction media features do teachers value the most?
\end{itemize}

@@

This question is mainly answered in the second survey, which, as indicated previously, was designed after initial analysis of the first. The kind of interaction is essential to design proper chatbots, or to deploy a technology instead of others; it is related to the perceived usefulness in a technology acceptance model. Essentially, if some new chatbot application does not have what teachers perceive as useful for their purposes, they are unlikely to adopt it. So knowing these opinions is quite important, and is the reason why we included it in the second survey.

The teachers were asked about what kind of features of the messaging system they value the most; the frequencies and percentages of their responses are given in Table \ref{tab:interaction}.

\begin{table}[h!]
\begin{tabular}{lcccc}
\hline
Interaction features & \multicolumn{2}{c}{Yes} & \multicolumn{2}{c}{No} \\
\hline
                    &  Freq. & \% &  Freq. & \% \\
                    \cline{2-5}
Analytics & 108 & 52.7 & 97 & 47.3 \\
Connectivity & 119 & 58.0 & 86 & 42.0 \\
Familiarity & 121 & 59.0 & 84 & 41.0 \\
Hidden Phone Number & 113 & 55.1 & 92 & 44.9 \\
Horizontally & 134 & 65.4 & 71 & 34.6 \\
Official formation & 42 & 20.5 & 163 & 79.5 \\
Pluggability & 65 & 31.7 & 140 & 68.3 \\
Sustainability & 157 & 76.6 & 48 & 23.4\\
Unidirectionality & 27 & 13.2 & 178 & 86.8 \\
Officiality & 150 & 73.2 & 55 & 26.8 \\
Synchrony & 45 & 22.0 & 160 & 78.0\\
Other & 5 & 2.4 & 200 & 97.6 \\
\hline
\end{tabular}
\vspace{0.8em}
\caption{Interaction media features valued by teachers. Details about each of these features can be read in Appendix I (Section \ref{app:questions}, second survey).}
\label{tab:interaction}
\end{table}

@@

The datasets generated and analysed during the current study are available in the {\tt anonymous} repository, accessible from https://anonymous.URL
